Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-26 article.tex
wordings

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is a high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough for being able to detect
interactions and, that it would therefore be preferrable to lower their
importance by changing appropriatly their magnitude. Additionnally, it is
well-known that neurons may also spike because of a global high activity
\citep{stetter2012model}. In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
\end{align}
which effect is to magnify the importance of spikes that occur in case of low
global activity (measured by $\sum_{j} x^{t}_j$), as observed for instance
around $t=4\text{s}$ in the figure. Note that the particular case where there
is no activity, i.e., $\sum_{j} x^{t}_j = 0$, is solved by setting $w(x^{t}_i)
= 1$.

@@

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
partial correlation coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$. %%  Partial
%% correlation can be interpreted in several ways. The partial
%% correlation coefficient $p_{i,j}$ can be shown to measure the
%% correlation between $X_i$ and $X_j$ when they are corrected for the
%% (linear) effect of all other variables in $X$ \cite{}.
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, ie. $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables. As an association measures between
neurons, it should thus naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks
\citep{de2004discovery,Schafer:2005}.
%% Given a set of multivariate Gaussian
%% random variables $X$, the (undirected) graph that connects all pairs
%% of variables $X_i$ and $X_j$ such that $p_{i,j}\neq 0$ is called a
%% Gaussian graphical models (GGMs). Because it measures conditional
%% dependencies, partial correlation should naturally only detect direct
%% associations between neurons. The superiority of partial correlation
%% over standard correlation as an association measure have already been
%% shown in the context of the inference of gene regulatory networks from
%% gene expression data \cite{}. We will compare both measures in Section
%% \ref{sec:results}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only affects slightly its
performances, with respect to the metric used in the Connectomics
challenge.

Practically, the computation of coefficient $p_{i,j}$ using Equation
\ref{eq:inverse} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that we have typically more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components \citep{bishop2006pattern} (with
$M=0.8 p$ in our experiments).

@@

We report here experiments on the \textit{normal}-1,2,3, and 4
datasets provided by the organizers of the Connectomics
challenge. Each of these datasets is obtained from the simulation
\citep{stetter2012model} of different neural networks of 1000 neurons
and about 15000 edges each (ie., a network density of about
1.5\%). Each neuron is described by a calcium fluorescence time series
of length $T=179500$. All inference methods compared here provide a
ranking of all pairs of neurons according to some association
score. To assess the quality of this ranking, we compute both ROC and
precision-recall curves against the ground truth network, which are
summarized using the area under the curve, denoted respectively AUROC
and AUPRC. Only the AUROC score was used to rank the challenge
participants but the precision-recall curve has been shown to be a
more sensible metric for network inference, especially when network
density is small (see e.g. \cite{schrynemackers2013protocols}).
Since neurons are not self-connected in the ground truth networks
(ie., $(i, i) \not \in E, \forall i \in V$), we have manually set the
score of such edges to the minimal possible association score before
computing ROC and PR curves.

>>>>

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is a high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough for being able to detect
interactions, and that it would therefore be preferrable to lower their
importance by changing appropriatly their magnitude. Additionnally, it is
well-known that neurons may also spike because of a global high activity
\citep{stetter2012model}. In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
\end{align}
which effect is to magnify the importance of spikes that occur in case of low
global activity (measured by $\sum_{j} x^{t}_j$), as observed for instance
around $t=4\text{s}$ in the figure. Note that the particular case where there
is no activity, i.e., $\sum_{j} x^{t}_j = 0$, is solved by setting $w(x^{t}_i)
= 1$.

@@

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
\textit{Partial correlation} coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$. %%  Partial
%% correlation can be interpreted in several ways. The partial
%% correlation coefficient $p_{i,j}$ can be shown to measure the
%% correlation between $X_i$ and $X_j$ when they are corrected for the
%% (linear) effect of all other variables in $X$ \cite{}.
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, i.e., $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables. As an association measure between
neurons, it should thus naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks
\citep{de2004discovery,Schafer:2005}.
%% Given a set of multivariate Gaussian
%% random variables $X$, the (undirected) graph that connects all pairs
%% of variables $X_i$ and $X_j$ such that $p_{i,j}\neq 0$ is called a
%% Gaussian graphical models (GGMs). Because it measures conditional
%% dependencies, partial correlation should naturally only detect direct
%% associations between neurons. The superiority of partial correlation
%% over standard correlation as an association measure have already been
%% shown in the context of the inference of gene regulatory networks from
%% gene expression data \cite{}. We will compare both measures in Section
%% \ref{sec:results}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only affects slightly its
performance, with respect to the metric used in the Connectomics
challenge.

Practically, the computation of all $p_{i,j}$ coefficients using Equation
\ref{eq:inverse} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that we have typically more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components \citep{bishop2006pattern} (with
$M=0.8 p$ in our experiments).

@@

We report here experiments on the \textit{normal}-1,2,3, and 4
datasets provided by the organizers of the Connectomics
challenge. Each of these datasets is obtained from the simulation
\citep{stetter2012model} of different neural networks of 1000 neurons
and about 15000 edges each (i.e., a network density of about
1.5\%). Each neuron is described by a calcium fluorescence time series
of length $T=179500$. All inference methods compared here provide a
ranking of all pairs of neurons according to some association
score. To assess the quality of this ranking, we compute both ROC and
precision-recall curves against the ground truth network, which are
summarized using the area under the curve, denoted respectively AUROC
and AUPRC. Only the AUROC score was used to rank the challenge
participants but the precision-recall curve has been shown to be a
more sensible metric for network inference, especially when network
density is small (see e.g. \cite{schrynemackers2013protocols}).
Since neurons are not self-connected in the ground truth networks
(i.e., $(i, i) \not \in E, \forall i \in V$), we have manually set the
score of such edges to the minimal possible association score before
computing ROC and PR curves.
